Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-08-28 manuscript/appB.Rmd
small edits from PBA on appendices

Results from statistical models
-------------------------------
*Tables begin on next page...*

>>>>

Results from statistical models
-------------------------------
In all tables, "mean_value" is the mean parameter estimate, "sd_value" is the standard deviation of the estimate from the MCMC, "lo_BCI" is the lower limit of the 95% Bayesian Credible Interval, and "up_BCI" is the upper limit of the 95% Bayesian Credible Interval

### 2015-08-28 manuscript/manuscript.Rmd
small edits from PBA on appendices

Results
-------
### Comparison of forecast models
The IPM had lower overall error (MAE, mean absolute error) compared to the QBM for three species (Table 1). The IPM MAE is significantly lower at $\alpha=0.05$ for *B. gracilis* (*P* = `r mae_ps[1,"value"]`), *H. comata* (*P* = `r mae_ps[2,"value"]`), and *P. smithii* (*P* = `r mae_ps[3,"value"]`). MAEs are statisticially similar between models for *P. secunda* (*P* = `r mae_ps[4,"value"]`). *P* values are highly sensitive to sample size, so not entirely appropriate in simulation exercises where we control the samples size. But, for our purposes they serve as relatively unbiased comparison metrics. In no case did the QBM significantly outperform the IPM (Table 1). The IPM was consistently more precise than the QBM, with lower distances between the 90% quantiles across all species (Table 1). In general the IPM outperformed the QBM because it had (1) lower MAE for three of the four species, (2) statistically similar MAE for the one other species, and (3) considerably more precise forecasts for all species.

@@

```{r get_results, echo=FALSE, eval=FALSE}
# setwd("~/Repos/MicroMesoForecast/analysis/")
# 
# ####
# ##  Mean parameter runs
# ####
# meanparam_ipm <- data.frame(cover=NA, species=NA, 
#                              climsim=NA, model=NA)
# ##  IPM 
# path <- "./ipm/simulations/results/climchange_meanparams/"
# ipmfiles_mean <- list.files(path)
# ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
# for(file in ipmfiles_mean){
#   tmp <- read.csv(paste(path,file,sep=""))
#   tmp$model <- "IPM"
#   meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
# }
# meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
# 
# newd <- data.frame(cover=NA, species=NA, 
#                    climsim=NA, model=NA, obscover=NA)
# for(dospp in unique(meanparam_ipm$species)){
#   tmp <- subset(meanparam_ipm, species==dospp)
#   obscover <- subset(tmp, climsim=="obs")["cover"]
#   tmp2 <- subset(tmp, climsim!="obs")
#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
#   newd <- rbind(newd, tmp2)
# }
# ipm_meansims <- newd[2:nrow(newd),]
# ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
# 
# ##  QBM
# meanparam_qbm <- data.frame(cover=NA, species=NA, 
#                              climsim=NA, model=NA)
# path <- "./quadBM/simulations/results/climatechange_meanparams/"
# qbmfiles_mean <- list.files(path)
# for(file in qbmfiles_mean){
#   tmp <- readRDS(paste(path,file,sep=""))
#   tmp$model <- "QBM"
#   meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
# }
# meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
# 
# newd <- data.frame(cover=NA, species=NA, 
#                    climsim=NA, model=NA, obscover=NA)
# for(dospp in unique(meanparam_qbm$species)){
#   tmp <- subset(meanparam_qbm, species==dospp)
#   obscover <- subset(tmp, climsim=="obs")["cover"]
#   tmp2 <- subset(tmp, climsim!="obs")
#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
#   newd <- rbind(newd, tmp2)
# }
# qbm_meansims <- newd[2:nrow(newd),]
# qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
# 
# ##  Merge IPM and QBM mean parameter results
# bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
# meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
#                        meandiff = mean(propdiff),
#                        hidiff = quantile(propdiff, probs = 0.95),
#                        lodiff = quantile(propdiff, probs = 0.05))
# setwd("~/Repos/MicroMesoForecast/manuscript")
```

@@

##  IPM results
# Read in data
ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
# Remove predictions >1 for fair comparison to [0,1] truncated quad model
resets <- which(ipm_onestep[,"cover.t1"]>1)
ipm_onestep[resets, "cover.t1"] <- 1
# Average predictions over quad-year reps
avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
                 avg_prediction = median(cover.t1),
                 observation = mean(cover.t0),
                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
# Calculate error
avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
# Aggregate over species
ipmstats <- ddply(avg_ipm, .(species), summarise,
                  mae = mean(abs(error)),
                  mean_cover = mean(observation*100),
                  quant_dist = mean(quant_dist))
ipmstats$model <- "IPM"

@@

```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates. For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates. These simulations use mean parameter values for clarity.", fig.width=4.5, fig.height=8, cache=FALSE, results='hide'}
setwd("~/Repos/MicroMesoForecast/analysis/")
#Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM using mean parameter values. Top panels show the mean predicted proportional change in cover. Lower panels show the
####
##  Mean parameter runs
####
# meanparam_ipm <- data.frame(cover=NA, species=NA, 
#                              climsim=NA, model=NA)
# ##  IPM 
# path <- "./ipm/simulations/results/climchange_meanparams/"
# ipmfiles_mean <- list.files(path)
# ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
# for(file in ipmfiles_mean){
#   tmp <- read.csv(paste(path,file,sep=""))
#   tmp$model <- "IPM"
#   meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
# }
# meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
# 
# newd <- data.frame(cover=NA, species=NA, 
#                    climsim=NA, model=NA, obscover=NA)
# for(dospp in unique(meanparam_ipm$species)){
#   tmp <- subset(meanparam_ipm, species==dospp)
#   obscover <- subset(tmp, climsim=="obs")["cover"]
#   tmp2 <- subset(tmp, climsim!="obs")
#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
#   newd <- rbind(newd, tmp2)
# }
# ipm_meansims <- newd[2:nrow(newd),]
# ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
# 
# ##  QBM
# meanparam_qbm <- data.frame(cover=NA, species=NA, 
#                              climsim=NA, model=NA)
# path <- "./quadBM/simulations/results/climatechange_meanparams/"
# qbmfiles_mean <- list.files(path)
# for(file in qbmfiles_mean){
#   tmp <- readRDS(paste(path,file,sep=""))
#   tmp$model <- "QBM"
#   meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
# }
# meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
# 
# newd <- data.frame(cover=NA, species=NA, 
#                    climsim=NA, model=NA, obscover=NA)
# for(dospp in unique(meanparam_qbm$species)){
#   tmp <- subset(meanparam_qbm, species==dospp)
#   obscover <- subset(tmp, climsim=="obs")["cover"]
#   tmp2 <- subset(tmp, climsim!="obs")
#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
#   newd <- rbind(newd, tmp2)
# }
# qbm_meansims <- newd[2:nrow(newd),]
# qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
# 
# ##  Merge IPM and QBM mean parameter results
# bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
# meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
#                        meandiff = mean(propdiff),
#                        hidiff = quantile(propdiff, probs = 0.95),
#                        lodiff = quantile(propdiff, probs = 0.05))
# dgd = position_dodge(width = 0.9)
# meanplot <- ggplot(meanparam_agg)+
#   geom_bar(aes(x=climsim, y=meandiff, fill=model),
#              position=dgd, stat="identity")+
#   theme_bw()+
#   scale_fill_manual(values=c("black", "grey45"), name="Model")+
#   facet_wrap("species", scales="free", nrow=1)+
#   scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
#   ylab("Mean proportional difference")+
#   xlab("Climate simulation")+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

equilibrium_cover <- subset(equilibrium_cover, vital!="A")
bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
equilibrium_cover[bothid,"climsim"] <- "zppttemp"
# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
sensplot <- ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
                              shape=vital, group=vital, linetype=vital))+
  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
  geom_line()+
  geom_point(size=4)+
  facet_wrap("species", scales = "free", ncol=1)+
  scale_linetype_discrete(name="Vital Rate")+
  scale_shape_discrete(name="Vital Rate")+
  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
  ylab("Equilibrium Cover (%)")+
  xlab("Simulation")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r figure_6, dependson="plot-options", fig.cap="Mean (points) and 90% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis. We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where "perturbed" and "observed" refer to the climate time series used to drive interannual variation in the simulations. Model error and parameter uncertainty were propagated through the simulation phase. Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide'}
####
##  Vary parameter runs
####
setwd("~/Repos/MicroMesoForecast/analysis/")
meanparam_ipm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, model=NA)
##  IPM 
path <- "./ipm/simulations/results/climchange_varyparams/"
ipmfiles_mean <- list.files(path)
ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
torm <- c(grep("10", ipmfiles_mean), grep("20", ipmfiles_mean), grep("30", ipmfiles_mean))
ipmfiles_mean <- ipmfiles_mean[-torm]
for(file in ipmfiles_mean){
  tmp <- read.csv(paste(path,file,sep=""))
  tmp$model <- "IPM"
  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
}
meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]

>>>>

Results
-------
```{r pvalues, echo=FALSE, eval=TRUE}
##  IPM results
# Read in data
setwd("/Users/atredenn/Repos/MicroMesoForecast/manuscript/")
ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
# Remove predictions >1 for fair comparison to [0,1] truncated quad model
resets <- which(ipm_onestep[,"cover.t1"]>1)
ipm_onestep[resets, "cover.t1"] <- 1
# Average predictions over quad-year reps
avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
                 avg_prediction = median(cover.t1),
                 observation = mean(cover.t0),
                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
# Calculate error
avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
# Aggregate over species
ipmstats <- ddply(avg_ipm, .(species), summarise,
                  mae = mean(abs(error)),
                  mean_cover = mean(observation*100),
                  quant_dist = mean(quant_dist))
ipmstats$model <- "IPM"

mae_ps <- melt(mae_pvals)
```

### Comparison of forecast models
The IPM had lower overall error (MAE, mean absolute error) compared to the QBM for three species (Table 1). The IPM MAE is significantly lower at $\alpha=0.05$ for *B. gracilis* (*P* = `r mae_ps[1,"value"]`), *H. comata* (*P* = `r mae_ps[2,"value"]`), and *P. smithii* (*P* = `r mae_ps[3,"value"]`). MAEs are statisticially similar between models for *P. secunda* (*P* = `r mae_ps[4,"value"]`). *P* values are highly sensitive to sample size, so not entirely appropriate in simulation exercises where we control the samples size. But, for our purposes they serve as relatively unbiased comparison metrics. In no case did the QBM significantly outperform the IPM (Table 1). The IPM was consistently more precise than the QBM, with lower distances between the 90% quantiles across all species (Table 1). In general the IPM outperformed the QBM because it had (1) lower MAE for three of the four species, (2) statistically similar MAE for the one other species, and (3) considerably more precise forecasts for all species.

@@

```{r get_results, echo=FALSE, eval=FALSE}
setwd("~/Repos/MicroMesoForecast/analysis/")

####
##  Mean parameter runs
####
meanparam_ipm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, model=NA)
##  IPM 
path <- "./ipm/simulations/results/climchange_meanparams/"
ipmfiles_mean <- list.files(path)
ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
for(file in ipmfiles_mean){
  tmp <- read.csv(paste(path,file,sep=""))
  tmp$model <- "IPM"
  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
}
meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]

newd <- data.frame(cover=NA, species=NA, 
                   climsim=NA, model=NA, obscover=NA)
for(dospp in unique(meanparam_ipm$species)){
  tmp <- subset(meanparam_ipm, species==dospp)
  obscover <- subset(tmp, climsim=="obs")["cover"]
  tmp2 <- subset(tmp, climsim!="obs")
  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
  newd <- rbind(newd, tmp2)
}
ipm_meansims <- newd[2:nrow(newd),]
ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))

##  QBM
meanparam_qbm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, model=NA)
path <- "./quadBM/simulations/results/climatechange_meanparams/"
qbmfiles_mean <- list.files(path)
for(file in qbmfiles_mean){
  tmp <- readRDS(paste(path,file,sep=""))
  tmp$model <- "QBM"
  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
}
meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]

##  Merge IPM and QBM mean parameter results
bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
                       meandiff = mean(propdiff),
                       hidiff = quantile(propdiff, probs = 0.95),
                       lodiff = quantile(propdiff, probs = 0.05))
setwd("~/Repos/MicroMesoForecast/manuscript")
```

@@

##  IPM results
# Read in data
setwd("/Users/atredenn/Repos/MicroMesoForecast/manuscript/")
ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
# Remove predictions >1 for fair comparison to [0,1] truncated quad model
resets <- which(ipm_onestep[,"cover.t1"]>1)
ipm_onestep[resets, "cover.t1"] <- 1
# Average predictions over quad-year reps
avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
                 avg_prediction = median(cover.t1),
                 observation = mean(cover.t0),
                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
# Calculate error
avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
# Aggregate over species
ipmstats <- ddply(avg_ipm, .(species), summarise,
                  mae = mean(abs(error)),
                  mean_cover = mean(observation*100),
                  quant_dist = mean(quant_dist))
ipmstats$model <- "IPM"

@@

```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates. For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates. These simulations use mean parameter values for clarity.", fig.width=8, fig.height=6, cache=FALSE, results='hide'}
setwd("~/Repos/MicroMesoForecast/analysis/")

@@

equilibrium_cover <- subset(equilibrium_cover, vital!="A")
bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
equilibrium_cover[bothid,"climsim"] <- "zppttemp"
# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
sensplot <- ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
                              shape=vital, group=vital, linetype=vital))+
  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
  geom_line()+
  geom_point(size=4)+
  facet_wrap("species", scales = "free", ncol=2)+
  scale_linetype_discrete(name="Vital Rate")+
  scale_shape_discrete(name="Vital Rate")+
  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
  ylab("Equilibrium Cover (%)")+
  xlab("Simulation")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r figure_6, dependson="plot-options", fig.cap="Mean (points) and 90% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis. We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where 'perturbed' and 'observed' refer to the climate time series used to drive interannual variation in the simulations. Model error and parameter uncertainty were propagated through the simulation phase. Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide'}
####
##  Vary parameter runs
####
setwd("~/Repos/MicroMesoForecast/analysis/")
meanparam_ipm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, model=NA)
##  IPM 
path <- "./ipm/simulations/results/climchange_varyparams/"
ipmfiles_mean <- list.files(path)
ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
torm <- c(grep("10", ipmfiles_mean), grep("20", ipmfiles_mean), grep("30", ipmfiles_mean))
ipmfiles_mean <- ipmfiles_mean[-torm]
for(file in ipmfiles_mean){
  tmp <- read.csv(paste(path,file,sep=""))
  tmp$model <- "IPM"
  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
}
meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
